Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-15 results.tex
typo in autoarima ref
getting autoarima citations to work
finished draft of conclusion

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the $MASE$ value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.

Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of these 45
predictions for each system: that is, the lowest error over all 15
trials and all three methods for each of the eight programs.  The WPE
is plotted against the corresponding $MASE$ value here in order to bring
out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    ($MASE$) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
The best-case prediction error for each of these eight systems is
roughly proportional to the weighted permutation entropy, which is
consistent with our first conjecture.  The dashed line in the figure,
a linear least-squares fit to these best-case errors\footnote{with the
  exception of {\tt dgesdd$_1$}, for reasons described below} captures
this rough proportionality.  This is not a formal result.  The three
methods used here were chosen to span the space of standard prediction
strategies, but they do not cover that space exhaustively.  Our goal
here is an empirical assessment of predictability and complexity, not
formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower $MASE$ values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
in the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail, including the measures taken to assure
meaningful comparisons across methods, trials, and programs, and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. $MASE$ plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons.
% 
% ; their $MASE$ scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in $MASE$ scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the
$MASE$ variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots.  The mean $MASE$ scores of
predictions generated with the nonlinear LMA method are generally
closer to the dashed line; the ARIMA method clusters are more
widely spread, the \naive ~clusters even more so.  This geometric
validation of the second conjecture in this paper is discussed in more
details in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the left-hand column of
Figure~\ref{fig:forecast-example} that the \naive ~and ARIMA
prediction methods do not perform as well on this signal as the
nonlinear LMA method.  The $MASE$ scores of the \naive ~and ARIMA
predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
across all 15 trials.  That is, these methods are performing only
$\approx 1.7$ times better than the random-walk method, a primitive
strategy that simply uses the current value as the prediction.
However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor $MASE$ scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  {\sl This indicates that these
  methods are not leveraging the available information in the signal.}
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the $MASE$ score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{auto-arima}.  If
a KPSS test of the time series in question indicates that it is
nonstationary, an integration term is added to the model.  This test
gives mixed results in the case of the \col process, flagging five of
the 15 trials as stationary and ten as nonstationary.  ARIMA models
without an integration term perform more poorly on this signal, which
increases the error, thereby spreading out the points\footnote{We
  tested this hypothesis by forcing the inclusion of an integration
  term in the five cases where a KPSS test indicated that such a term
  was not needed.  This action removed the spread, pushing all 15 of
  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
  tight cluster.}.

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the available
information: methods that use linear models, for instance, may fail
when applied to nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
This match between model and task is an issue in the \svdfive
experiments as well.  The $MASE$ scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~$MASE$ score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. $MASE$ plot---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the corresponding method is able to
leverage, and that a different method might do better.  Indeed, for
\svdfive, the LMA method produces a $MASE$ score of $ 0.718\pm 0.048 $
and a cluster of results that is much closer to the dashed line on the
WPE-$MASE$ plot.  Again, this validates our second conjecture: the LMA
method can capture and reproduce the way in which the \svdfive system
processes information, but the \naive ~and ARIMA prediction methods
cannot.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well on this
signal, yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714
\pm 0.075$ (ARIMA).  This pushes the corresponding clusters of points
in the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well
above the trend followed by the other seven signals.  The $MASE$
scores of the predictions that were produced by the \naive ~method for
this system, however, are highly inconsistent.  The majority of the
blue diamond-shaped points on the top plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a $MASE$ score of
0.6.  In five of the 15 \svdone trials, however, there were step
changes in the signal.  The \naive ~method has a very difficult time
with signals like this, particularly if there are multiple step
changes.  This raised the $MASE$ scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near $MASE \approx 2.5$, as well as two points that
  are beyond the domain of the graph, at $MASE$ 11.2 and 14.8.}, and
in turn raising both the mean and variance of this set of trials.

The effects described in the previous paragraph are exacerbated by the
way $MASE$ is calculated.  Recall that $MASE$ scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the $MASE$ score, this effect can shift points
leftwards on a WPE vs. $MASE$ plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated $MASE$ scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone $MASE$ values.  For all of these reasons, we view \svdone
as an outlier and exclude it from the fit calculation of the dashed
line in Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.

Figure~\ref{fig:wpe_vs_mase_best} shows the {\sl best} of these 45
predictions for each system: that is, the lowest error over all 15
trials and all three methods for each of the eight programs.  The WPE
is plotted against the corresponding MASE value here in order to bring
out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
The best-case prediction error for each of these eight systems is
roughly proportional to the weighted permutation entropy, which is
consistent with our first conjecture.  The dashed line in the figure,
a linear least-squares fit to these best-case errors\footnote{with the
  exception of {\tt dgesdd$_1$}, for reasons described below} captures
this rough proportionality.  This is not a formal result.  The three
methods used here were chosen to span the space of standard prediction
strategies, but they do not cover that space exhaustively.  Our goal
here is an empirical assessment of predictability and complexity, not
formal results about a ``best'' predictor for a given time series.
There may be other methods that produce lower MASE values than those
in Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
in the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail, including the measures taken to assure
meaningful comparisons across methods, trials, and programs, and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems. \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons.
% 
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots.  The mean MASE scores of
predictions generated with the nonlinear LMA method are generally
closer to the dashed line; the ARIMA method clusters are more
widely spread, the \naive ~clusters even more so.  This geometric
validation of the second conjecture in this paper is discussed in more
details in the following paragraphs.

Though \col is a very simple program, its dynamics are actually quite
complicated, as discussed in Section~\ref{sec:intro}.  Recall from the
graphical representations in the left-hand column of
Figure~\ref{fig:forecast-example} that the \naive ~and ARIMA
prediction methods do not perform as well on this signal as the
nonlinear LMA method.  The MASE scores of the \naive ~and ARIMA
predictions are $0.571 \pm 0.002$ and $0.599 \pm 0.211$, respectively,
across all 15 trials.  That is, these methods are performing only
$\approx 1.7$ times better than the random-walk method, a primitive
strategy that simply uses the current value as the prediction.
However, the WPE value for the \col trials is $0.513 \pm 0.003$, which
is in the center of the complexity spectrum described in
Section~\ref{sec:intro}.

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

@@

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{autoARIMA}.  If
a KPSS test of the time series in question indicates that it is
nonstationary, an integration term is added to the model.  This test
gives mixed results in the case of the \col process, flagging five of
the 15 trials as stationary and ten as nonstationary.  ARIMA models
without an integration term perform more poorly on this signal, which
increases the error, thereby spreading out the points\footnote{We
  tested this hypothesis by forcing the inclusion of an integration
  term in the five cases where a KPSS test indicated that such a term
  was not needed.  This action removed the spread, pushing all 15 of
  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
  tight cluster.}.

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the available
information: methods that use linear models, for instance, may fail
when applied to nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the corresponding method is able to
leverage, and that a different method might do better.  Indeed, for
\svdfive, the LMA method produces a MASE score of $ 0.718\pm 0.048 $
and a cluster of results that is much closer to the dashed line on the
WPE-MASE plot.  Again, this validates our second conjecture: the LMA
method can capture and reproduce the way in which the \svdfive system
processes information, but the \naive ~and ARIMA prediction methods
cannot.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well on this
signal, yielding MASE scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
0.075$ (ARIMA).  This pushes the corresponding clusters of points in
the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
the trend followed by the other seven signals.  The MASE scores of the
predictions that were produced by the \naive ~method for this system,
however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6.  In five of the 15 \svdone trials, however, there were step
changes in the signal.  The \naive ~method has a very difficult time
with signals like this, particularly if there are multiple step
changes.  This raised the MASE scores of these trials, pushing the
corresponding points to the right\footnote{This includes the cluster
  of three points near MASE $\approx 2.5$, as well as two points that
  are beyond the domain of the graph, at MASE 11.2 and 14.8.}, and in
turn raising both the mean and variance of this set of trials.

The effects described in the previous paragraph are exacerbated by the
way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone
as an outlier and exclude it from the fit calculation of the dashed
line in Figure~\ref{fig:wpe_vs_mase_best}.
